Allen Developing Human Brain 450K Data Normalization
========================================================
## By: Sumaiya Islam
## Date updated: August 10, 2016
  
### Script contents:
  - detection and correction for technical batch variation using PCA and ComBat, respectively, of post-mortem human brain samples analyzed by Illumina HM450K platform from Allen Brain Institute (http://download.alleninstitute.org/brainspan/Methylation/). 
   
### A. Set up working directory & packages

R version 3.2.3 (2015-12-10)

We will initially set our working directory and load our libraries.

```{r,include=FALSE}
library(reshape)
library(ggplot2)
library(RColorBrewer)
library(grid)
library(gridExtra)
library(sva)
library(rama)
library(methylumi)
library(gplots)
library(marray)
library(lumi)
library(lattice)
library("RColorBrewer")
library(knitr)
library(xtable)
library(wateRmelon)
library(limma)
library(sva)
library(RPMM)
```

## Heat scree plot Function
```{r}
### Function of association meta variable with PC (ANOVA)
heat_scree_plot<-function(Loadings, Importance, Num, Order){
  adjust<-1-Importance[1]
  pca_adjusted<-Importance[2:length(Importance)]/adjust
  pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted)))
  
  scree<-ggplot(pca_df[which(pca_df$PC<Num),],aes(PC,adjusted_variance))+geom_bar(stat = "identity",color="black",fill="grey")+theme_bw()+
        theme(axis.text = element_text(size =12),
              axis.title = element_text(size =15),
              plot.margin=unit(c(1,1.5,0.2,2.25),"cm"))+ylab("Variance")+
    scale_x_continuous(breaks = seq(1,Num,1))
  
  #### Heat
  ## correlate meta with PCS
  ## Run anova of each PC on each meta data variable

  aov_PC_meta<-lapply(1:ncol(meta_categorical), function(covar) sapply(1:ncol(Loadings), function(PC) summary(aov(Loadings[,PC]~meta_categorical[,covar]))[[1]]$"Pr(>F)"[1]))
  cor_PC_meta<-lapply(1:ncol(meta_continuous), function(covar) sapply(1:ncol(Loadings), function(PC) (cor.test(Loadings[,PC],as.numeric(meta_continuous[,covar]),alternative = "two.sided", method="spearman", na.action=na.omit, exact=FALSE)$p.value)))
  names(aov_PC_meta)<-colnames(meta_categorical)
  names(cor_PC_meta)<-colnames(meta_continuous)
  aov_PC_meta<-do.call(rbind, aov_PC_meta)
  cor_PC_meta<-do.call(rbind, cor_PC_meta)
 aov_PC_meta<-rbind(aov_PC_meta, cor_PC_meta)
  aov_PC_meta<-as.data.frame(aov_PC_meta)
  #adjust
  aov_PC_meta_adjust<-aov_PC_meta[,2:ncol(aov_PC_meta)]
  
    
  #reshape
  avo<-aov_PC_meta_adjust[,1:(Num-1)]
  avo_heat_num<-apply(avo,2, as.numeric)
  avo_heat<-as.data.frame(avo_heat_num)
  colnames(avo_heat)<-sapply(1:(Num-1), function(x) paste("PC",x, sep=""))
  avo_heat$meta<-rownames(avo)
  avo_heat_melt<-melt(avo_heat, id=c("meta"))
  
  # cluster meta data
  ord <- Order
  meta_var_order<-unique(avo_heat_melt$meta)[rev(ord)]
  avo_heat_melt$meta <- factor(avo_heat_melt$meta, levels = meta_var_order)
  
  # color if sig
   avo_heat_melt$Pvalue<-sapply(1:nrow(avo_heat_melt), function(x) if(avo_heat_melt$value[x]>=0.9){">=0.9"}else{
    if(avo_heat_melt$value[x]>=0.5){">=0.5"}else{
      if(avo_heat_melt$value[x]>=0.1){">=0.1"}else{"<0.1"}}})
  avo_heat_melt$Pvalue<-sapply(1:nrow(avo_heat_melt), function(x) if(avo_heat_melt$value[x]<=0.001){"<=0.001"}else{
     if(avo_heat_melt$value[x]<=0.01){"<=0.01"}else{
       if(avo_heat_melt$value[x]<=0.05){"<=0.05"}else{">0.05"}}})
  
  heat<-ggplot(avo_heat_melt, aes(variable,meta, fill = Pvalue)) +
  geom_tile(color = "black",size=0.5) +
  theme_gray(8)+scale_fill_manual(values=c("#084594","#4292c6","#9ecae1","#deebf7"))+
      theme(axis.text = element_text(size =10, color="black"),
            axis.text.x = element_text(),
          axis.title = element_text(size =15),
          legend.text = element_text(size =14),
          legend.title = element_text(size =12),
          legend.position = c(1, 0), legend.justification = c(1,0),
          plot.margin=unit(c(0,2.25,1,1),"cm"))+
    xlab("Principal Component")+ylab(NULL)
  
  grid.arrange(scree, heat, ncol=1)
}
```


### B. Load files

#### We will be analyzing the normalized Allen Brain datasets

```{r}

load("Allen_brain_BMIQ.RData") # normalized beta values
load("Allen_Dev_Brain_450K_Corrected.RData") # cell-type corrected beta values
load("Allen_Human_Developing_Brain_450K_Meta_Data.RData") # associated meta data
cell.proportions<-read.csv("Allen_CETS_Neuronal_Glia_Proportions.csv", header=TRUE, row.names=1) # predicted neuron and glial cell proportions based on CETS
head(cell.proportions)

# check for NAs in data
ind<-is.row.na(Allen_brain_BMIQ) # The function returns a vector of logical variables, one for each row of the matrix. The variable is TRUE if the row does not contain any missing values and FAlSE otherwise.
length(na.count<-which(ind=="FALSE")) # 0 (there are no rows that contain a NA in the raw data)


uncor.dat<-Allen_brain_BMIQ
cor.dat<-Allen.cell.corrected
meta<-Allen.Dev.Brain.meta.fin
identical(rownames(cell.proportions), as.character(meta$Sample_ID)) # TRUE
meta$Neuron<-as.numeric(cell.proportions$neuron)
meta$Glia<-as.numeric(cell.proportions$glia)
str(meta)
```


## PCA Scree Heatmap for uncorrected data

```{r warning=FALSE, fig.height=9, fig.width=11}
## PCA
PCA_full<-princomp(uncor.dat[complete.cases(uncor.dat),]) # scaling is not necessary for normalized dataset
Loadings<-as.data.frame(unclass(PCA_full$loadings))
vars <- PCA_full$sdev^2
Importance<-vars/sum(vars)
adjust<-1-Importance[1]
pca_adjusted<-Importance[2:length(Importance)]/adjust
(pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted))))


#Specify which covariates are categorical and/or categorical
colnames(meta)
meta_categorical<-meta[,c("Plate", "Sentrix_ID", "Row", "Brain_Region")]  # input column numbers in meta that contain categorical variables
meta_continuous<-meta[,c("Neuron", "Age_Months")] # input column numbers in meta that contain continuous variables
meta_continuous<-data.frame(meta_continuous)

# Specify the number of PCs you want shown (usually # of samples in the dataset)
Num<-20

# Designate what order you want the variables to appear (continuous variables rbinded to categorical variables in function)
Order<-c(4,5,6,3,2,1)

#Apply function on PCA results, pulls in the meta data and beta values from above
heat_scree_plot(Loadings, Importance, Num, Order)
```


We can observe that there are some significant Batch, Chip and Row effects that I will attempt to correct. However, they occur in the same PCs as Brain_Region and Age_Months making it difficult to ComBat correct---will seek help from the group on this. 

## PCA Scree Heatmap for normalized data

```{r warning=FALSE, fig.height=9, fig.width=11}
## PCA
PCA_full<-princomp(cor.dat[complete.cases(cor.dat),]) # scaling is not necessary for normalized dataset
Loadings<-as.data.frame(unclass(PCA_full$loadings))
vars <- PCA_full$sdev^2
Importance<-vars/sum(vars)
adjust<-1-Importance[1]
pca_adjusted<-Importance[2:length(Importance)]/adjust
(pca_df<-data.frame(adjusted_variance=pca_adjusted, PC=seq(1:length(pca_adjusted))))


#Specify which covariates are categorical and/or categorical
colnames(meta)
meta_categorical<-meta[,c("Plate", "Sentrix_ID", "Row", "Brain_Region")]  # input column numbers in meta that contain categorical variables
meta_continuous<-meta[,c("Neuron", "Age_Months")] # input column numbers in meta that contain continuous variables
# meta_continuous<-data.frame(meta_continuous)

# Specify the number of PCs you want shown (usually # of samples in the dataset)
Num<-20

# Designate what order you want the variables to appear (continuous variables rbinded to categorical variables in function)
Order<-c(4,5,6,3,2,1)

#Apply function on PCA results, pulls in the meta data and beta values from above
heat_scree_plot(Loadings, Importance, Num, Order)
```

