Allen Developing Human Brain 450K Data Normalization
========================================================
## By: Sumaiya Islam
## Date updated: August 10, 2016
  
### Script contents:
  - normalization of post-mortem human brain samples analyzed by Illumina HM450K platform from Allen Brain Institute (http://download.alleninstitute.org/brainspan/Methylation/). 
   
### A. Set up working directory & packages

R version 3.2.3 (2015-12-10)

We will initially set our working directory and load our libraries.
```{r CARE_Tissue_setup, include = FALSE, echo=FALSE}
setwd("/home/sschaffner/Allen_Brain")
library(methylumi)
library(gplots)
library(marray)
library(lumi)
library(lattice)
library("RColorBrewer")
library(knitr)
library(xtable)
library(wateRmelon)
library(limma)
library(sva)
library(RPMM)
```

### B. Load files


```{r CARE_Tissue_fileLoad, echo=FALSE}
load(file = "Allen_Human_Developing_Brain_450K_Filtered_Data.RData") # filtered Allen data
load(file="Allen_Human_Developing_Brain_450K_Meta_Data.RData")
load("Priest.RData") # Priest methylumi object for fData
dim(Allen.filtered)
dim(Priest)
```

NOTE: Will need to go back and perform quantro and quantile normalization on the data!!!

### C. Quantro

The quantro R-package can be used to test for global differences between
groups of distributions which assess whether global normalization methods
such as quantile normalization should be applied.

```{r}
library(quantro)
Allen_brain_betas<-as.matrix(Allen.filtered)
head(Allen.Dev.Brain.meta.fin)
str(Allen.Dev.Brain.meta.fin)
matboxplot(Allen_brain_betas, groupFactor = Allen.Dev.Brain.meta.fin$Brain_Region,brewer.n=16, xaxt = "n",main = "Beta Values")
```

We can see differences in the median methylation signals between different brain regions.

```{r}
library(doParallel)
registerDoParallel(cores=4)
qtestPerm <- quantro(Allen_brain_betas, Allen.Dev.Brain.meta.fin$Brain_Region, B = 1000)
qtestPerm
quantroPlot(qtestPerm)
#p<0.05 then there are global differences in distributions across groups
```

We can see that the quantro stat (922.91056) is significant (p = 0),
indicating that there are global differences in distributions across groups
so quantile normalization is not appropriate for this dataset.


### D. Probe-type Normalization: BMIQ

The 450K Illumina Infinium Array has inherent variation associated with its methodologies which must be accounted for in our analyses. Much of this variation is attributed to the use of two types of probes used in the array, Type I and Type II.

Type I probes contain two bead types corresponding to an unmethylated (U) or methylated (M) status. Type I probes obtain methylation status as a result of fluoresence expressed after a single base pair extension occurs just after the target basepair, resulting in only one color channel being utilized (red). Type I probes also assume that any CpG sites underlying the probe are of the same status as the query site (methylated or unmethylated). The beta values for Type I probes are then determined by this formula b= M/(U + M). Importantly, Type I probes are enriched in regions of high CpG density (carry 3 or more CpG sites underlying the probe body), particularly those associated with promoters). Type II probes tend to occur in lower CpG density regions (carry 3 or less CpG sites underlying the probe body). Type II probes do not assume the methylation status of underlying CpG sites within the probe and so consist of a combination of degenerate probes with varying combinations of up to three underlying CpG sites. Type II probes also detect methylation status with a single base pair extension except that the site being extended is the CpG site of detection and so require two fluorescent colors green for methylated (M) and red for unmethylated (U) sites. Type II probe beta values are calculated using this formula b = Green (M)/(Red (U) + Green (M)). In terms of considerations for normalization, Type I probes have a much higher dynamic range than Type II probes, which may introduce an enrichment bias towards Type I probes when ranking probes in supervised analyses (Teschendorff et al 2013, Bioinformatics). 

Due to these inherent differences between Type I and Type II probes used in the Illumina Infinium 450K array several groups in the field have deveolped various normalization analyses to correct for the differences between these probes. We will be using an intra-sample normalization method that corrects for probe-type differences called BMIQ (Beta MIxture Quantile dilation) (Teschendorff et al 2013, Bioinformatics).

#### BMIQ (Beta MIxture Quantile dilation)

BMIQ is an intra-sample normalisation procedure, correcting the bias of type-2 probe values. BMIQ uses a 3-step procedure: (i) fitting of a 3-state beta mixture model, (ii) transformation of state-membership probabilities of type2 probes into quantiles of the type1 distribution, and (iii) a conformal transformation for the hemi-methylated probes. Exact details can be found in the reference (Teschendorff et al 2013, Bioinformatics).


```{r CARE_Tissue_BMIQ, echo=FALSE}

dim(Priest.filtered<-Priest.filtered[featureNames(Priest.filtered)%in%rownames(Allen.filtered),])
head(probe_design<-as.character(fData(Priest.filtered)$INFINIUM_DESIGN_TYPE))
probe_design.v<- replace(probe_design, probe_design=="I", 1)
probe_design.cor<- replace(probe_design.v, probe_design.v=="II", 2)
probe_design.cor<-as.numeric(probe_design.cor)
identical(nrow(Allen.filtered), length(probe_design.cor))

# Run BMIQ across each sample
betas_normalized<-apply(Allen.filtered, 2, function(x) BMIQ(x,probe_design.cor)) # this code takes a long time---best to run it overnight esp if you have a lot of samples

# extract normalized beta values and reshape
betas_normalized_betas<-lapply(1:length(betas_normalized), function(x) betas_normalized[[x]]$nbeta)
betas_normalized_betas<-do.call(rbind, betas_normalized_betas)
betas_norm.fin<-t(betas_normalized_betas)
colnames(betas_norm.fin)<-colnames(Allen_brain_betas)
head(betas_norm.fin)


# save normalized data
Allen_brain_BMIQ <- betas_norm.fin
save(Allen_brain_BMIQ, file = "Allen_brain_BMIQ.RData")
```

## Comparing raw and normalized datasets

For all samples (PBMC samples are in yellow while buccal samples are in blue):
```{r CARE_Tissue_compareAll, echo=FALSE}
par(mfrow = c(1, 2))
plot(density(Allen_brain_betas),legend=NULL, lty = 1, xlim = c(-6, 6), ylim=c(0,0.2), main = "Raw", col=2)
plot(density(Allen_brain_BMIQ),legend=NULL, lty = 1, xlim = c(-6, 6), ylim=c(0,0.2), main = "BMIQ", col=3)

## use ggplot2 to plot density plots of raw and normalized data
# extract beta matrix for raw and normalized datasets

# methylumi object - betas.raw<-betas(Allen_brain_betas)
# methylumi object - betas.norm<-betas(Allen_brain_BMIQ)
betas.raw <- Allen_brain_betas
betas.norm <- Allen_brain_BMIQ

# randomly sample 10,000 probes from the datasets (same random probes in each dataset)
random.probes<-sample(1:nrow(betas.raw), 10000)
betas.raw.subset<-betas.raw[random.probes,]
betas.norm.subset<-betas.norm[random.probes,]
# melt each dataset
head(betas.raw.melt<-melt(betas.raw.subset))
head(betas.norm.melt<-melt(betas.norm.subset))
# remove NAs
betas.raw.melt.clean<-betas.raw.melt[which(!(is.na(betas.raw.melt$value))),]
betas.norm.melt.clean<-betas.norm.melt[which(!(is.na(betas.norm.melt$value))),]
# add descriptor for each datatype and add meta data to each dataset
betas.raw.melt.clean$Data<-"Raw"
betas.norm.melt.clean$Data<-"Normalized"
head(betas.raw.melt.clean)

# methylumi - meta.raw<-pData(Allen_brain_betas)
# methylumi - meta.norm<-pData(Allen_brain_BMIQ)
# methylumi - meta.raw$SampleName<-rownames(meta.raw)
# methylumi - meta.norm$SampleName<-rownames(meta.norm)

meta.raw <- Allen.Dev.Brain.meta.fin
meta.norm <- Allen.Dev.Brain.meta.fin
# meta.raw$SampleName <- Allen.Dev.Brain.meta.fin$Sample_ID
# meta.norm$SampleName <- Allen.Dev.Brain.meta.fin$Sample_ID

head(betas.raw.plot<-merge(betas.raw.melt.clean, meta.raw, by.x="X2", by.y="Sample_ID"))
head(betas.norm.plot<-merge(betas.norm.melt.clean, meta.norm, by.x="X2", by.y="Sample_ID"))
# combine both datasets
betas.plot<-rbind(betas.raw.plot, betas.norm.plot)
#re-order levels of Data variable
betas.plot$Data<-factor(betas.plot$Data, levels=c("Raw", "Normalized"))

# assign a colour to each brain region sample
region_colours <- matrix(c("DFC", "#C0392B",
"VFC", "#E74C3C",
"MFC", "#9B59B6",
"OFC", "#8E44AD",
"M1C", "#2980B9",
"S1C", "#3498DB",
"IPC", "#1ABC9C",
"A1C", "#16A085",
"STC", "#27AE60",
"ITC", "#2ECC71",
"V1C", "#F1C40F",
"HIP", "#F39C12",
"AMY", "#E67E22",
"STR", "#D35400",
"MD", "#BDC3C7",
"CBC", "#95A5A6"),ncol=2,byrow=TRUE)
colnames(region_colours) <- c("Brain_Region","Region_colour")
as.table(region_colours)
head(region_colours)

# merge metadata with colours list, by brain region 
head(meta.colours<-merge(betas.plot, region_colours, by.y="Brain_Region"))
colour_list <- unique(meta.colours$Region_colour)
colour_list <- as.character(colour_list[1:length(colour_list)])
table(unique(meta.colours$Region_colour))
dim(meta.colours)

# plot density plots
ggplot(meta.colours, aes(value, group = X2, colour = as.factor(Region_colour))) + geom_density() + theme_bw() + facet_wrap(~Data) 
```

We can see that there is reduced sample-to-sample variability in the BMIQ-normalized dataset versus the raw data. 


### F. Plot beta values before/after normalization

We will show the distribution of beta values for the entire dataset (all samples normalized together) in its raw form and final normalized form.

```{r CARE_Tissue_betaPlot, echo=FALSE} 
type <- fData(Priest.filtered)$INFINIUM_DESIGN_TYPE

par(mfrow = c(1, 1))
plot(c(0, 1) , c(0, 10), type = "n", xlab = "Beta values", ylab = "Density")
lines(density(na.omit(betas.raw)[type=="I",]),col="blue",lty=1)
lines(density(na.omit(betas.raw)[type=="II",]),col="blue",lty=2, lwd=1.5)
lines(density(na.omit(betas.norm)[type=="I",]),col="red",lty=1)
lines(density(na.omit(betas.norm)[type=="II",]),col="red",lty=2, lwd=1.5)
legend("top", inset=.05, cex=0.6, c("Raw Type 1","Raw Type 2", "Normalized Type 1", "Normalized Type 2"), col=c("blue","blue","red", "red"), lty=c(1,2,1,2), horiz=FALSE)
```

We can see that there is improved peak-to-peak overlap of the Type 1 and Type 2 probes with the final normalized dataset as desired (ie Type 1 and 2 probes have more similar dynamic ranges for their beta values). 

We will continue on with our analysis using the BMIQ-normalized datasets.