---
title: "SCRIPT for Hilary to run on all probes"
author: "Cassia Warren"
date: "March 29, 2017"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


#source("https://bioconductor.org/biocLite.R")
#biocLite("limma")
#biocLite("edgeR")
#install.packages('dplyr')
#biocLite("ComplexHeatmap")
#biocLite("gplots")
#install.packages(NMF)
#biocLite("yeast2.db")
#install.packages('VennDiagram')
library(VennDiagram)
library(yeast2.db) 	
library(limma)
library(edgeR)
library(ComplexHeatmap)
library(ggplot2)
library(reshape2)
library(tidyverse)
library(mvtnorm)
library(gplots)
library(gridExtra)
library(plyr)
library(dplyr)
library(NMF)
library(lattice)
```


**Load the data and check that the columns are in the same order**
^^ repeat with the Dasen normalized data and the New BMIQ data

```{r load}

load("/Volumes/Lexar/New corrected/GSE43414_batch_cor.RData")

load("/Volumes/Lexar/New corrected/GSE43414_cell_cor.RData")

load("/Volumes/Lexar/New corrected/Meta_batch_cor.RData")

```

Lisas code to rearrange
^^ repeat with the Dasen normalized data and the New BMIQ data
```{r rea}

## add another broad region column
meta$broad_regions <- ifelse(meta$Tissue == "cerebellum", "cerebellum","cortex")
meta$tissue_color <- lapply(meta$Tissue,function(x){
  if (x == "cerebellum") {y <- "red"}
  if (x == "frontal cortex") {y <- "blue"}
  if (x == "superior temporal gyrus") {y <- "orange"}
  if (x == "entorhinal cortex") {y <- "yellow"}
  y
  })

meta$broad_colors <- lapply(meta$broad_regions,function(x){
  if (x == "cerebellum") {y <- "red"}
  if (x == "cortex") {y <- "blue"}
  y
})

## transpose data such that probe names are colnames, and rows are patient samples
transpose_GSE43414_cell_cor <- t(GSE43414_cell_cor)
## order metadata by brain region
#data has barcode NOT gsm as column names
meta_order_by_brain_regions <- meta %>% arrange(Tissue)
matches_GSE43414_cell_cor <- match(meta_order_by_brain_regions$barcode, rownames(transpose_GSE43414_cell_cor))

GSE43414_cell_cor_sorted_by_brain_regions <- t(transpose_GSE43414_cell_cor[matches_GSE43414_cell_cor,])
#repeat for batch

transpose_GSE43414_batch_cor <- t(GSE43414_batch_cor)
## order metadata by brain region
GSE43414_batch_cor_sorted_by_brain_regions <- t(transpose_GSE43414_batch_cor[matches_GSE43414_cell_cor,])

#Check for identical 
identical(colnames(GSE43414_batch_cor_sorted_by_brain_regions),colnames(GSE43414_cell_cor_sorted_by_brain_regions)) # TRUE

identical(colnames(GSE43414_cell_cor_sorted_by_brain_regions),as.character(meta_order_by_brain_regions$barcode)) #TRUE


```


PLOTS _ repeat with both data sets (change "GSE43414" for the data set names to the other data set name)
```{r plots}

# make lables
brain_regions <- meta_order_by_brain_regions$broad_regions
tissue <- meta_order_by_brain_regions$Tissue

# no sample rearange
aheatmap(GSE43414_batch_cor_sorted_by_brain_regions, Colv = NA,annCol = list(Brain_Region = brain_regions, Tissue = tissue), main =  "Uncorrected heatmap")

#dendogram of samples
aheatmap(GSE43414_batch_cor_sorted_by_brain_regions, annCol = list(Brain_Region = brain_regions, Tissue = tissue), main =  "Uncorrected heatmap")

# no sample rearange
aheatmap(GSE43414_cell_cor_sorted_by_brain_regions, Colv = NA, annCol = list(Brain_Region = brain_regions, Tissue = tissue, main = "Corrected heatmap"))

#dendogram of samples
aheatmap(GSE43414_cell_cor_sorted_by_brain_regions, annCol = list(Brain_Region = brain_regions, Tissue = tissue, main = "Corrected heatmap"))

```


**Want to see what thier expression is and if it changes between corrected and non corercted data sets. **
Repeat with both data sets
```{r graph}

# create a combined data set
batch_sub <- GSE43414_batch_cor_sorted_by_brain_regions

cell_sub <- GSE43414_cell_cor_sorted_by_brain_regions

```

Repeat with both data sets
```{r correlation}
#sample to sample correlations based on brain region


aheatmap(cor(batch_sub),Colv = NA, Rowv = NA,  annCol = list(Brain_Region = brain_regions, Tissue = tissue), annRow = list(Brain_Region = brain_regions, Tissue = tissue),main = "uncorrected sample to sample correlation of 100 probes")

aheatmap(cor(cell_sub),Colv = NA, Rowv = NA,  annCol = list(Brain_Region = brain_regions, Tissue = tissue), annRow = list(Brain_Region = brain_regions, Tissue = tissue), main = "corrected sample to sample correlation of 100 probes")