---
title: "Basic Heatmaps (BMIQ)"
output: 
  html_document: 
    keep_md: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)


#source("https://bioconductor.org/biocLite.R")
#biocLite("limma")
#biocLite("edgeR")
#install.packages('dplyr')
#biocLite("ComplexHeatmap")
library(NMF)
library(dplyr)
```


```{r load}

load("~/Methylhomies/GSE43414_batch_cor.RData")

load("~/Methylhomies/GSE43414_cell_cor.RData")

load("~/Methylhomies/Meta_batch_cor.RData")

```

Remove Braak stage exludes and NAs

```{r remove}
meta2 <- na.omit(meta) #remove NA
meta <- meta2[!c(meta2$braak.stage=="Exclude"),] #remove exlucdes
```

Lisas code to rearrange

```{r rea}

## add another broad region column
meta$broad_regions <- ifelse(meta$Tissue == "cerebellum", "cerebellum","cortex")
meta$tissue_color <- lapply(meta$Tissue,function(x){
  if (x == "cerebellum") {y <- "red"}
  if (x == "frontal cortex") {y <- "blue"}
  if (x == "superior temporal gyrus") {y <- "orange"}
  if (x == "entorhinal cortex") {y <- "yellow"}
  y
  })

meta$broad_colors <- lapply(meta$broad_regions,function(x){
  if (x == "cerebellum") {y <- "red"}
  if (x == "cortex") {y <- "blue"}
  y
})

## transpose data such that probe names are colnames, and rows are patient samples
transpose_GSE43414_cell_cor <- t(GSE43414_cell_cor)
## order metadata by brain region
#data has barcode NOT gsm as column names
meta_order_by_brain_regions <- meta %>% arrange(Tissue)
matches_GSE43414_cell_cor <- match(meta_order_by_brain_regions$barcode, rownames(transpose_GSE43414_cell_cor))

GSE43414_cell_cor_sorted_by_brain_regions <- t(transpose_GSE43414_cell_cor[matches_GSE43414_cell_cor,])
#repeat for batch

transpose_GSE43414_batch_cor <- t(GSE43414_batch_cor)
## order metadata by brain region
GSE43414_batch_cor_sorted_by_brain_regions <- t(transpose_GSE43414_batch_cor[matches_GSE43414_cell_cor,])

#Check for identical 
identical(colnames(GSE43414_batch_cor_sorted_by_brain_regions),colnames(GSE43414_cell_cor_sorted_by_brain_regions)) # TRUE

identical(colnames(GSE43414_cell_cor_sorted_by_brain_regions),as.character(meta_order_by_brain_regions$barcode)) #TRUE


```


```{r}
brain_regions <- meta_order_by_brain_regions$broad_regions
tissue <- meta_order_by_brain_regions$Tissue

set.seed(1)
probes <- sample(rownames(GSE43414_cell_cor_sorted_by_brain_regions),1000)

aheatmap(GSE43414_batch_cor_sorted_by_brain_regions[probes,], Colv = TRUE,annCol = list(Brain_Region = brain_regions, Tissue = tissue), main =  "Batch", width = 2, height = 2)

aheatmap(GSE43414_cell_cor_sorted_by_brain_regions[probes,], Colv = TRUE, annCol = list(Brain_Region = brain_regions, Tissue = tissue), main = "Cell Type Corrected", width = 2, height = 2)
```

```{r graph}

# create a combined data set
batch_sub <- GSE43414_batch_cor_sorted_by_brain_regions

cell_sub <- GSE43414_cell_cor_sorted_by_brain_regions

```

```{r correlation}
#sample to sample correlations based on brain region

aheatmap(cor(batch_sub),Colv = NA, Rowv = NA,  annCol = list(Brain_Region = brain_regions, Tissue = tissue), annRow = list(Brain_Region = brain_regions, Tissue = tissue),main = "Non-Cell Type Corrected BMIQ")

aheatmap(cor(cell_sub),Colv = NA, Rowv = NA,  annCol = list(Brain_Region = brain_regions, Tissue = tissue), annRow = list(Brain_Region = brain_regions, Tissue = tissue), main = "Cell Type Corrected BMIQ")